---
title: "An Illustration of the Central Limit Theorem"
author: "Andrew Reynolds"
date: "September 22, 2015"
output: pdf_document
---

## Overview

If $X$ is an exponential random variable with mean $\mu_X$ and variance $\sigma_X^2$, and $\bar{X}$ is the mean of an iid sample of size $n$, then the Central Limit Theorem says that as $n$ becomes large, $\bar{X}$ becomes approximately normal, with mean and variance equal to 
    $$\mu_{\bar{X}} = \mu_{X} \quad \text{and} \quad 
    \sigma_{\bar{X}}^2 = \dfrac{\sigma_X^2}{n}$$
respectively.

In this project, we aim to verify the above claims by simulation, using the `rexp()` function. We let $n = 40$ and take a large sample (1000 independent draws) from $\bar{X}$. If the Central Limit Theorem is true, then our sample ought to be approximately normal, with mean and variance approximately equal to the figures above (with $n = 40$). We will check this graphically and numerically.

## Simulations

We let $X$ be an exponential random variable with $\lambda = 0.2$, so that $\mu_X = \dfrac{1}{\lambda} = 5$ and $\sigma_X = \dfrac{1}{\lambda} = 5$. The following code simulates a single sample of size $n = 40$ and computes its mean. Note that `rexp()` encodes $\lambda$ as `rate`.

```{r}
set.seed(100)
example <- rexp(40, rate = 0.2)
mean(example)
```

This mean is a single draw from $\bar{X}$. To obtain our sample which should represent $\bar{X}$, we repeat the above code 1000 times.

```{r}
sample <- numeric(0)
for (i in 1:1000) {
    sample <- c(sample, mean(rexp(40, rate = 0.2)))
}
```

## Sample Mean versus Theoretical Mean:

We plot the histogram of our sample and mark the sample mean with a vertical blue line, and the theoretical value with a red line.

```{r fig.height = 3.25}
hist(sample, main = "Histogram of sample - means indicated")
segments(x0 = mean(sample), y0 = 150, y1 = 300, lwd = 2, col = "blue")
segments(x0 = 5, y0 = 150, y1 = 0, lwd = 2, col = "red")
```

This figure shows that our sample mean is very close to the theoretical mean of 5. Let's check this numerically:

```{r}
mean(sample) - 5
```

The difference is small.

## Sample Variance versus Theoretical Variance:

Lastly, we compare the variance of our sample with theoretical value of $\dfrac{25}{40}$.

```{r}
sd(sample)^2 - 25/40
```

We see that the approximation is a good one. We now include another histogram of our sample with the mean indicated (we'll use 5, since there is no visible difference between 5 and the sample mean) as well as one standard deviation from the mean in both directions. Again, blue represents the measured standard deviation and red represents the theoretical value.

```{r fig.height = 3.25}
hist(sample, main = "Histogram of sample - sd's indicated")
abline(v = 5, lwd = 2)
segments(x0 = 5 + c(-1,1)*sd(sample), y0 = 150, y1 = 300, lwd = 2, col = "blue")
segments(x0 = 5 - sd(sample), x1 = 5 + sd(sample), y0 = 200, col = "blue")
segments(x0 = 5 + c(-1,1)*5/sqrt(40), y0 = 150, y1 = 0, lwd = 2, col = "red")
segments(x0 = 5 - 5/sqrt(40), x1 = 5 + 5/sqrt(40), y0 = 75, col = "red")
```

## Distribution

Now we overlay the histogram of our sample with the a scaled version of the normal bell curve with mean and standard deviation specified by the Central Limit Theorem. Note that the height of the `dnorm` curve is multiplied by 500 to scale appropriately (the total height of the rectangles is 1000 and they have width 0.5, and so the total area is 500).

```{r fig.height = 3.25}
hist(sample, main = "Histogram of sample with theoretical (normal) curve")
curve(500*dnorm(x, mean = 5, sd = 5/sqrt(40)), 
      col = "darkblue", lwd = 2, add = TRUE)
```

The histogram matches the theoretical curve very well, with a peak near the mean and roughly symmetric tails on either side.
